---
title: "Docker-instructions"
author: "Thomas Rauter"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Docker-instructions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{=html}
<style>
.sourceCode pre {
  background-color: #f7f7f7;
  padding: 10px;
  border: 1px solid #ccc;
  border-radius: 5px;
  overflow-x: auto;
  position: relative;
}

.copy-btn {
  position: absolute;
  top: 10px;
  right: 10px;
  border: none;
  background: #4CAF50;
  color: white;
  padding: 5px 10px;
  cursor: pointer;
  border-radius: 5px;
}

.copy-btn:hover {
  background: #45a049;
}
</style>
```
```{=html}
<script>
document.addEventListener("DOMContentLoaded", function() {
  document.querySelectorAll(".sourceCode pre").forEach(function(block) {
    var button = document.createElement("button");
    button.innerHTML = "Copy";
    button.className = "copy-btn";
    block.style.position = "relative"; // Ensure button is correctly positioned
    block.appendChild(button);

    button.addEventListener("click", function() {
      // Select only the code inside the block and not the button
      var code = block.querySelector('code').innerText;
      navigator.clipboard.writeText(code).then(function() {
        button.innerHTML = "Copied!";
        setTimeout(function() {
          button.innerHTML = "Copy";
        }, 2000);
      });
    });
  });
});
</script>
```
# Pulling the `Docker` Container

Pull `Docker` container:

```{r eval=FALSE}
docker pull thomasrauter/splineomics:0.1.0
```

If you face 'permission denied' issues, check out [this vignette](https://raw.githubusercontent.com/csbg/SplineOmics/main/doc/Docker_permission_denied.html)

# Running the `Docker` Container

To run the `Docker` container, you can use one of the following commands, depending on your operating system. Before running the command, ensure that you are in a directory containing two subfolders: `input` and `output`. These will be used to transfer files between your local machine and the Docker container.

For Linux and macOS:

```{r eval=FALSE}
# Bash
docker run -it -d \
    -v $(pwd)/input:/home/rstudio/input \
    -v $(pwd)/output:/home/rstudio/output \
    -p 8888:8787 \
    -e PASSWORD=123 \
    --name splineomics \
    thomasrauter/splineomics:0.1.0
```

For Windows:

```{r eval=FALSE}
# PowerShell
docker run -it -d `
    -v "${PWD}\input:/home/rstudio/input" `
    -v "${PWD}\output:/home/rstudio/output" `
    -p 8888:8787 `
    -e PASSWORD=123 `
    --name splineomics `
    thomasrauter/splineomics:0.1.0
```

Once the container is running, open a web browser and navigate to <http://localhost:8888>. Log in using the following credentials:

-   Username: rstudio

-   Password: The one you set with the -e PASSWORD=123 option (123 in this case)

As long as the container is running, you can work on that localhost page with RStudio, where also the `SplineOmics` package is installed. `/home/rstudio/` is your R session working folder.

Stop the container:

```{r eval=FALSE}
docker stop splineomics
```

Start the container again:

```{r eval=FALSE}
docker start splineomics
```

# Input and Output File Management

The `input` and `output` directories on your local machine are mounted to corresponding directories inside the Docker container. This allows seamless file transfer between your local machine and the container.

-   Place your input files (e.g., data, metadata, annotation files) in the `input` directory on your local machine. These files will automatically appear in `/home/rstudio/input` inside the container.

-   Any files generated by RStudio within the container should be saved to `/home/rstudio/output`. These files will automatically appear in the `output` directory on your local machine.

| **Local Directory** | **Docker Container Directory** | **Description**                                                |
|-------------------|--------------------|----------------------------------|
|                     | `/home/rstudio/`               | Working directory in the RStudio session inside the container. |
| `input/`            | `/home/rstudio/input/`         | Place your input files here on your local machine.             |
| `output/`           | `/home/rstudio/output/`        | Files generated in the container will appear here locally.     |

# Inspect `Docker` container installations

To see all the R packages and system installations that make up the `Docker` container, you can run the following command in the terminal of RStudio on your localhost browser page.

```{r eval=FALSE}
cp -r /log home/rstudio/output
```

Because the `/home/rstudio/output` dir is mounted to your local filesystem, this will make the installation log files available there.

# Installing additional R packages in the container

New R packages can be installed the normal way:

```{r eval=FALSE}
install.packages("package_name")
```

However, note that any packages installed in the running container will be lost if the container is deleted or rebuilt.

# Permanent additions

If you want to permanently add R packages, R scripts, or other files to the SplineOmics `Docker` image, you can use it as a base image for building a new image. This will ensure that all changes are saved into the new image, rather than being lost when the container is deleted.

For example:

```{r eval=FALSE}
# Use the SplineOmics image as the base image
FROM thomasrauter/splineomics:0.1.0

# Install the data.table package permanently
RUN R -e "install.packages('data.table')"

# Optionally, add custom R scripts to the image
COPY your_script.R /home/rstudio/your_script.R

# Set the working directory
WORKDIR /home/rstudio

# Expose RStudio Server port
EXPOSE 8787

# Start RStudio server
CMD ["/init"]

# Build new image:
# docker build -t your_new_image_name .
```

Run the container of the new image with the commands described above.
