---
title: "Docker-instructions"
author: "Thomas Rauter"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Docker-instructions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Pulling the `Docker` Container

Pull `Docker` container:

```{r eval=FALSE}
# Docker Hub repo still private, will make it public then
docker pull thomasrauter/splineomics:0.1.0
```

If you face 'permission denied' issues, check out [this vignette](https://raw.githubusercontent.com/csbg/SplineOmics/main/doc/Docker_permission_denied.html)

# Running the `Docker` Container

To run the `Docker` container, you can use one of the following commands, depending on your operating System. The respective command needs to be run in a folder where the subfolders `input` and `output` exist. Place your data and meta (and annotation) files in `input`, and receive your output from the container in the `output` folder.

For Linux and macOS:

```{r eval=FALSE}
# Bash
docker run -it -d \
    -v $(pwd)/input:/home/rstudio/input \
    -v $(pwd)/output:/home/rstudio/output \
    -p 8888:8787 \
    -e PASSWORD=123 \
    --name splineomics \
    thomasrauter/splineomics:0.1.0
```

For Windows:

```{r eval=FALSE}
# PowerShell
docker run -it -d `
    -v "${PWD}\input:/home/rstudio/input" `
    -v "${PWD}\output:/home/rstudio/output" `
    -p 8888:8787 `
    -e PASSWORD=123 `
    --name splineomics `
    thomasrauter/splineomics:0.1.0
```

Once the container is running, open a web browser and navigate to <http://localhost:8888>. Use rstudio as the username and the password you set with the -e PASSWORD=123 option.

As long as the container is running, you can work on that localhost page with RStudio, where also the `SplineOmics` package is installed. `/home/rstudio/` is your R session working folder.

Direct all generated results to `/home/rstudio/output`, which is mounted to your local folder `output`. Your results will be there.

Stop the container:

```{r eval=FALSE}
docker stop splineomics
```

Start the container again:

```{r eval=FALSE}
docker start splineomics
```

# Inspect Docker container installations

To see all the R packages and system installations that make up the Docker container, you can run the following command in the terminal of RStudio on your localhost browser page.

```{r eval=FALSE}
cp -r /log home/rstudio/output
```

Because the `/home/rstudio/output` dir is mounted to your local filesystem, this will make the installation log files available there.
