---
title: "FAQ"
author: "Thomas Rauter"
date: "30 Oktober, 2025"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{faq}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Frequently Asked Questions

--------------------------------------------------------------------------------

## FAQ on data pre-processing

### Does SplineOmics handle missing values?

Yes. SplineOmics can handle missing values gracefully because it uses 
`limma` as the statistical engine for model fitting and differential 
analysis. The `limma` package is designed to work with incomplete 
expression matrices — missing values are simply ignored when estimating 
model parameters and computing contrasts, without causing errors.

This means you do not necessarily need to impute or remove features with 
a few missing entries before running SplineOmics. However, if a feature 
has too many missing values (for example, across most samples), consider 
removing it.

--------------------------------------------------------------------------------

## FAQ on parameter selection

### Which design formula should I use?

The choice of design formula depends on your experimental setup and on 
whether you want to model your conditions independently or jointly.

- Isolated designs (no interaction between condition and time)  
  Use this when you want to analyse each condition completely separately, 
  for example, to get independent spline fits for each treatment.  
  In this case, `SplineOmics` will simply process each dataset 
  sequentially for convenience, so you do not need to run the code twice.  
  This approach is ideal when you do not expect conditions to share 
  information or influence one another.

- Integrated designs (conditions modelled jointly with a time interaction)  
  Use this when you want to fit all conditions in a single model and 
  allow their time trends to differ through an interaction term between 
  condition and time.  
  With this approach, datasets are modelled jointly and can borrow 
  statistical power from each other, leading to more stable variance 
  estimates and more sensitive detection of shared temporal patterns.  
  This also enables access to the full range of `limma` result categories, 
  including category 2 and category 3 results (see the vignette on 
  [limma result categories](limma_result_categories.html)).

### Should I use B-splines or natural cubic splines?

Both spline types can model smooth trends across time or another 
continuous variable, but they differ in how local their flexibility is.

- B-splines are locally adaptive: changing one knot affects the fitted 
  curve only in a small neighbourhood around that knot.  
  This makes them ideal when you expect local variations (for example, 
  short-term biological responses) and want the rest of the curve to 
  remain stable. The trade-off is that B-splines typically use more 
  degrees of freedom, so the model can become more complex.

- Natural cubic splines enforce global smoothness: each basis function 
  extends across the entire range, so adjusting one part of the curve 
  slightly influences the whole fit. They use fewer degrees of freedom 
  and can be more stable when you expect overall smooth behaviour, but 
  they are less suitable if local detail is important.

In short:  
> Choose B-splines when local flexibility matters, and natural cubic 
> splines when you want a smoother, more global trend.

### How many degrees of freedom should I use for the splines?

There is no single optimal choice — the best number of degrees of 
freedom (dof) depends on the smoothness and complexity of your data.  
In practice, 2 or 3 degrees of freedom work well in most cases.

Using more degrees of freedom makes the spline wigglier and increases 
the risk of overfitting, while using only 1 degree of freedom is usually 
too restrictive to capture meaningful trends.

If you set the degrees of freedom to 0, `SplineOmics` will automatically 
determine the optimal value using leave-one-out cross-validation, 
selecting the dof that provides the best predictive performance for your 
dataset.

### Should I use array weights?

Yes, in most cases this is recommended. Since `SplineOmics` builds on 
`limma`, it inherits support for array weights, which help to correct 
for heteroskedasticity (unequal variances) often present in time-series 
data.

By estimating a quality weight for each sample, `SplineOmics` can give 
less influence to noisier samples and more weight to consistent ones, 
resulting in smoother spline fits and higher statistical power.

### How many clusters should I use?

There is no universal rule for choosing the optimal number of clusters.  
Using more clusters makes each cluster purer — its centroid represents 
its members more precisely — but it also makes downstream interpretation 
more complex and fragmented.

To help with this choice, `SplineOmics` provides several aids:

- For each cluster, it reports the variance explained by the cluster 
  centroid, both as a mean value and as a distribution histogram, with 
  guidance on how to interpret these values.  
- For each individual feature, it shows how well it is represented by its 
  cluster centroid.  
- You can specify a range or custom set of cluster numbers (for example, 
  2–10 or {2, 5, 6}), and `SplineOmics` automatically selects the best 
  one using the Bayesian Information Criterion (BIC):

  \[
  \text{BIC} = n_{\text{obs}} 
  \log\left(\frac{\text{tot\_within}}{n_{\text{obs}}}\right) 
  + k \log(n_{\text{obs}}) \times p
  \]

The model with the lowest BIC is chosen as the optimal clustering 
configuration.

--------------------------------------------------------------------------------

## General FAQ

### Can SplineOmics handle datasets with more than two conditions?

Not directly. `SplineOmics` is designed for pairwise comparisons between 
conditions. You can, however, analyse experiments with more than two 
conditions by either:

- using an isolated design, where each condition is analysed separately, 
  or  
- performing all pairwise comparisons among the conditions of interest.

These approaches allow you to explore multiple conditions, but they do 
not integrate them into a single joint model. Integrated designs are 
currently limited to comparisons between two conditions at a time.

# Session Info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
