---
title: "Glycan-analysis"
author: "Thomas Rauter"
date: "13 February, 2024"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Glycan-analysis}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

# About this vignette

This tutorial intends to showcase and explain the capabilities of the
**SplineOmics** package by walking through a real and complete glycan analysis
example, from start to finish. **SplineOmics** is explained in more
detail in the **get-started** vignette, where a proteomics example is
covered. This vignette is more focused on showing how glycan data can
be used, and because of this, less details about the overall package are
provided here.

## Data Overview

This dataset originates from a **time-series glycan experiment**
designed to study Chinese Hamster Ovary (CHO) cells. The data is compositional,
meaning that the observed glycoform abundances represent parts of a whole and 
are constrained by a constant-sum constraint, making standard statistical
analyses inappropriate without appropriate log-ratio transformations. The
experiment
involved cultivating cells in **eight bioreactors**, with **four
bioreactors** subjected to a temperature shift after 146 hours
(experimental condition) and the remaining **four bioreactors**
maintained without a temperature shift (control condition).

### Timepoints

Samples were collected at **7 distinct time points** throughout the
experiment, specifically: `"120h"`, `"144h"`, `"168h"`, `"192h"`, `"216h"`,
`"288h"`, and `"336h"` after cultivation
start. Each time point was sampled from all eight bioreactors, but E17_336 is 
missing, therefore resulting in a total of **55 samples**.

### Effects in the Experiment: Reactor and Batch

In this experiment, there are two effects to consider: **Reactor** and 
**Batch**.

1. **Reactor**:
   - This refers to the different bioreactors used for cell cultivation, which
   can exhibit substantial variability.
   - Each reactor was assigned a single condition: either constant temperature
   or temperature-shifted. As a result,
   **condition and reactor are confounded**.
   - Reactor should not be treated as a fixed effect to simply remove its
   influence. Instead, it is treated as a **random effect**, which allows
   us to model its variability appropriately.

2. **Batch**:
   - This refers to the two separate glycan analysis batches.
   - Batch is considered a **batch effect** with respect to the condition 
   (constant temperature vs. temperature-shifted).

Since Condition and Reactor are confounded, the variability due to the reactors
cannot be directly separated from the condition. Instead,
**linear mixed models (LMMs)** are used to attribute the reactor as a random
effect, allowing us to account for its variability while isolating the effects
of the condition. This approach ensures that the analysis appropriately handles
the hierarchical structure of the data and avoids incorrect conclusions.

In this vignette, we will demonstrate how to use linear mixed models to
address these challenges and properly account for both reactor and plate
effects.

### Further info

The data matrix comprises **glycoforms as rows** (such as G0/G0 and none/G0F)
and
**samples as columns**, providing glycoform measurements for all time points.
Each
glycoform as a row stands for a combination of sugars (glycans) that can be
attached to the 
left and right side of the product antibody, that we produced in our CHO cell 
cultivation. For example G0/G0 means that the glycan G0 was attached to both 
sides, and none/G0F means at the left side, there was no glycan, and on the 
right side, there was the G0F glycan.

The goal of this experiment is to investigate the effect of a
temperature shift during CHO cell cultivation on the antibody glycan
dynamics over time.

Note: This is not the original dataset, as it has not yet been published
at the time of this vignette's creation. For demonstration purposes, the
glycans have been randomly shuffled, and 2% has been randomly added or 
substracted off each value.

## Analysis Goals

The main objectives of this analysis are:

-   **Identify glycans with significant temporal changes**: Among the
    glycoforms measured, the goal is to identify those that
    exhibit significant changes in abundance over time.

-   **Cluster glycans based on temporal patterns**: Glycoforms showing
    significant temporal changes (hits) will be grouped into clusters
    based on their time-dependent expression patterns.

-   **Assess the impact of temperature shifts on temporal patterns**:
    The analysis will determine whether the temporal patterns of glycoform
    abundance are affected by the temperature shift, i.e., whether glycoform
    abundance dynamics differ over time under temperature shift
    conditions compared to controls.

### Note

The documentation of all the **SplineOmics** package functions can be
viewed [here](https://csbg.github.io/SplineOmics/reference)

# Load the packages

```{r setup, eval = TRUE}
library(SplineOmics)
library(here) # For managing filepaths
library(dplyr) # For data manipulation
library(knitr) # For Showing the head of the data and the meta tables.
library(compositions) # For clr transforming the glycan data
```

# Load the files

```{r load the files, , eval = TRUE}
data <- read.csv(
    system.file(
        "extdata",
        "glycan_data.csv",
        package = "SplineOmics"
    ),
    stringsAsFactors = FALSE
)

meta <- read.csv(
    system.file(
        "extdata",
        "glycan_meta.csv",
        package = "SplineOmics"
    ),
    stringsAsFactors = FALSE
)

# Set the first column as row names and remove it from the dataframe
rownames(data) <- data[[1]] # Assign first column as row names
data <- data[, -1] # Remove the first column


# Make data to a numeric matrix (required by SplineOmics)
data <- data.matrix(data)
```

## Show top rows of data

```{r Show data head, eval = TRUE}
knitr::kable(
    head(data),
    format = "markdown"
)
```

## Show top rows of meta

```{r Show meta head, eval = TRUE}
knitr::kable(
    head(meta),
    format = "markdown"
)
```

# Perform EDA (exploratory data analysis)

```{r Load EDA arguments, eval = TRUE}
# Those fields are mandatory, because we believe that when such a report is
# opened after half a year, those infos can be very helpful.
report_info <- list(
    omics_data_type = "Glycan",
    data_description = "clr transformed timeseries fractional abundance
    glycoform data of CHO cells, batch corrected",
    data_collection_date = "September 2024",
    analyst_name = "Thomas Rauter",
    contact_info = "thomas.rauter@plus.ac.at",
    project_name = "DGTX"
)

report_dir <- here::here(
    "results",
    "explore_data"
)
```

Compositional data, such as glycan profiles expressed as relative abundances,
are inherently constrained by a constant sum and therefore reside in a simplex
rather than Euclidean space. This violates the assumptions of standard
statistical methods. To address this, we applied a centered log-ratio (CLR)
transformation, which appropriately maps the data into real space by accounting
for the compositional structure. While some studies apply a simple
log-transformation to stabilize variance, this approach ignores the relative
nature of the data and may lead to biased or misleading results. CLR
transformation is thus the more appropriate and statistically sound choice
for downstream modeling.

```{r clr transform the data, eval = TRUE}
# TRANSPOSE: make samples rows, features columns
data_t <- t(data)

# Handle zeros (simple pseudocount, crude but common)
data_t[data_t == 0] <- 1e-6

# Apply CLR
clr_data_t <- compositions::clr(data_t)

# Transpose back to original shape: features as rows, samples as columns
clr_data <- t(clr_data_t)
clr_data <- unclass(clr_data)
```

```{r Create the SplineOmics object, eval = TRUE}
# splineomics now contains the SplineOmics object.
splineomics <- SplineOmics::create_splineomics(
    data = clr_data,
    meta = meta,
    report_info = report_info,
    condition = "Condition", # Column of meta that contains the levels.
    meta_batch_column = "Batch" # For batch effect removal in the plots
)

# Special print.SplineOmics function leads to selective printing
print(splineomics)
```

```{r Run EDA function, eval = FALSE}
plots <- SplineOmics::explore_data(
    splineomics = splineomics, # SplineOmics object
    report_dir = report_dir
)
```

[Here](https://drive.google.com/uc?id=1h4aOSwxUHAp5KAoAYVOlL_IRotV4EtY0)
you can see the HTML report of the explore_data() function with the NOT
batch-corrected data, and
[here](https://drive.google.com/uc?id=1op3oCt7eNhrUaofIkAyNJrSqpschU78b)
the report for the batch-corrected data.

# Run limma spline analysis

In this example, we are skipping finding the best hyperparameters with
the screen_limma_hyperparams() function, because we already have a clear
idea of what do use.

Lets define our parameters and put them into the `SplineOmics` object:

```{r Update the SplineOmics object, eval = TRUE}
splineomics <- SplineOmics::update_splineomics(
    splineomics = splineomics,
    use_array_weights = FALSE,
    # Reactor as random effect.
    design = "~ 1 + Condition*Time + Batch + (1|Reactor)",
    mode = "integrated", # means limma uses the full data for each condition.
    spline_params = list(
        spline_type = c("n"), 
        dof = c(2L) 
    )
)
```

The term '(1|Reactor)' in the design formula are the random effects. Therefore,
the linear mixed models that we use will model Reactor as a random effect.

Run the `run_limma_splines()` function with the updated SplineOmics
object:

```{r limma spline analysis, eval = TRUE}
splineomics <- SplineOmics::run_limma_splines(
    splineomics = splineomics
)
```

Note that for our dataset, the Wilcoxon signed rank exact test revealed that we
have a problem of heteroscedasticity for our data. This means that the variance
between the samples is significantly higher in the stationary phase compared to
the exponential phase. This is a violation for the assumption of
homoscedasticity of linear models, which means that the derived p-values are not
trustworthy. To adress this issue, limma has a function (arrayWeights) which 
downweights samples with higher variance, which we then use here automatically.
Further, the eBayes function of limma with the robust = TRUE argument avoids
using variance "outliers" for the variance shrinkage.

# Build limma report

The topTables of all three limma result categories can be used to
generate p-value histograms an volcano plots.

```{r build limma report, eval = TRUE}
plots <- SplineOmics::create_limma_report(
    splineomics = splineomics,
    report_dir = withr::local_tempdir()
)
```

You can view the generated analysis report of the create_limma_report
function
[here](https://drive.google.com/uc?id=1jgE8d5Gs12xnMATaitXxp28j2nAKCX1Y).

# Cluster the hits (significant features)

Prepare arguments:

```{r prepare arguments, eval = TRUE}
nr_clusters <- list(
    constant = 2,
    tshifted = 2
)

plot_info <- list(
    y_axis_label = "Glycan fractional abundance",
    time_unit = "hours",
    treatment_labels = list(
        tshifted = "temp shift"
    ),
    treatment_timepoints = list(
        tshifted = 146
    )
)

# Those are not genes for this glycan analysis here, but this argument expects
# the feature names (which usually are gene names, which is why it is called
# like this.
genes <- rownames(data)

plot_options <- list(
    # When meta_replicate_column is not there, all datapoints are blue.
    meta_replicate_column = "Reactor", # Colors the data points based on Reactor
    cluster_heatmap_columns = FALSE
)
```

Run the clustering:

```{r cluster the hits, eval = TRUE}
clustering_results <- SplineOmics::cluster_hits(
    splineomics = splineomics,
    adj_pthresh_time_effect = 0.05,
    adj_pthresh_avrg_diff_conditions = 0.05,
    adj_pthresh_interaction_condition_time = 0.05,
    nr_clusters = nr_clusters,
    genes = genes
)
```

Generate the report:

```{r generate clustering report, eval = TRUE}
plots <- SplineOmics::create_clustering_report(
    report_payload = clustering_results$report_payload,
    plot_info = plot_info,
    plot_options = plot_options,
    verbose = TRUE,
    max_hit_number = 5,
    report_dir = withr::local_tempdir()
)
```

You can view the generated analysis report of the cluster_hits function
[here](https://drive.google.com/uc?id=1HXKoyOniO6-tDC-uCiGgpaQcjUQ9-zyl).

As discussed before, there are three limma result categories. The
cluster_hits() report shows the results of all three, if they are
present (category 2 and 3 can only be generated when the design formula
contains an interaction effect).

# Session Info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
