---
title: "Glycan-analysis"
author: "Thomas Rauter"
date: "13 February, 2024"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Glycan-analysis}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# About this vignette

This tutorial intends to showcase and explain the capabilities of the
**SplineOmics** package by walking through a real and complete glycan analysis
example, from start to finish. **SplineOmics** is explained in more
detail in the **get-started** vignette, where a proteomics example is
covered. This vignette is more focused on showing how glycan data can
be used, and because of this, less details about the overall package are
provided here.

## Data Overview

This dataset originates from a **time-series glycan experiment**
designed to study Chinese Hamster Ovary (CHO) cells. The experiment
involved cultivating cells in **eight bioreactors**, with **four
bioreactors** subjected to a temperature shift after 146 hours
(experimental condition) and the remaining **four bioreactors**
maintained without a temperature shift (control condition).

### Timepoints

Samples were collected at **7 distinct time points** throughout the
experiment, specifically: `"120h"`, `"144h"`, `"168h"`, `"192h"`, `"216h"`,
`"288h"`, and `"336h"` after cultivation
start. Each time point was sampled from all eight bioreactors, but E17_336 is 
missing, therefore resulting in a total of **55 samples**.

### Effects in the Experiment: Reactor and Batch

In this experiment, there are two effects to consider: **Reactor** and 
**Batch**.

1. **Reactor**:
   - This refers to the different bioreactors used for cell cultivation, which can exhibit substantial variability.
   - Each reactor was assigned a single condition: either constant temperature or temperature-shifted. As a result, **condition and reactor are confounded**.
   - Reactor should not be treated as a fixed effect to simply remove its influence. Instead, it is treated as a **random effect**, which allows us to model its variability appropriately.

2. **Batch**:
   - This refers to the two separate glycan analysis batches.
   - Batch is considered a **batch effect** with respect to the condition 
   (constant temperature vs. temperature-shifted).

Since Condition and Reactor are confounded, the variability due to the reactors cannot be directly separated from the condition. Instead, **linear mixed models (LMMs)** are used to attribute the reactor as a random effect, allowing us to account for its variability while isolating the effects of the condition. This approach ensures that the analysis appropriately handles the hierarchical structure of the data and avoids incorrect conclusions.

In this vignette, we will demonstrate how to use linear mixed models to address these challenges and properly account for both reactor and plate effects.

### Further info

The data matrix comprises **glycoforms as rows** (such as G0/G0 and none/G0F) and
**samples as columns**, providing glycoform measurements for all time points. Each
glycoform as a row stands for a combination of sugars (glycans) that can be attached to the 
left and right side of the product antibody, that we produced in our CHO cell 
cultivation. For example G0/G0 means that the glycan G0 was attached to both 
sides, and none/G0F means at the left side, there was no glycan, and on the 
right side, there was the G0F glycan.

The goal of this experiment is to investigate the effect of a
temperature shift during CHO cell cultivation on the antibody glycan
dynamics over time.

Note: This is not the original dataset, as it has not yet been published
at the time of this vignette's creation. For demonstration purposes, the
glycans have been randomly shuffled, and 2% has been randomly added or 
substracted off each value.

## Analysis Goals

The main objectives of this analysis are:

-   **Identify glycans with significant temporal changes**: Among the
    glycoforms measured, the goal is to identify those that
    exhibit significant changes in abundance over time.

-   **Cluster glycans based on temporal patterns**: Glycoforms showing
    significant temporal changes (hits) will be grouped into clusters
    based on their time-dependent expression patterns.

-   **Assess the impact of temperature shifts on temporal patterns**:
    The analysis will determine whether the temporal patterns of glycoform
    abundance are affected by the temperature shift, i.e., whether glycoform
    abundance dynamics differ over time under temperature shift
    conditions compared to controls.

### Note

The documentation of all the **SplineOmics** package functions can be
viewed [here](https://csbg.github.io/SplineOmics/reference)

# Load the packages

```{r setup, eval = TRUE}
# library(SplineOmics)
library(devtools)
devtools::load_all()
library(here) # For managing filepaths
library(dplyr) # For data manipulation
library(knitr) # For Showing the head of the data and the meta tables.
```

# Load the files

```{r load the files}
data <- read.csv(
  system.file(
    "extdata",
    "glycan_data.csv",
    package = "SplineOmics"
  ),
  stringsAsFactors = FALSE
)

meta <- read.csv(
  system.file(
    "extdata",
    "glycan_meta.csv",
    package = "SplineOmics"
  ),
  stringsAsFactors = FALSE
)

# Set the first column as row names and remove it from the dataframe
rownames(data) <- data[[1]]  # Assign first column as row names
data <- data[-1]  # Remove the first column


# Make data to a numeric matrix (required by SplineOmics)
data <- data.matrix(data)   
```

## Show top rows of data

```{r Show data head, eval = TRUE}
knitr::kable(
  head(data),
  format = "markdown"
  )
```

## Show top rows of meta

```{r Show meta head, eval = TRUE}
knitr::kable(
  head(meta),
  format = "markdown"
  )
```

# Perform EDA (exploratory data analysis)

```{r Load EDA arguments, eval = TRUE}
# Those fields are mandatory, because we believe that when such a report is
# opened after half a year, those infos can be very helpful.
report_info <- list(
  omics_data_type = "Glycan",
  data_description = "Glycoform data of CHO cells",
  data_collection_date = "September 2024",
  analyst_name = "Thomas Rauter",
  contact_info = "thomas.rauter@plus.ac.at",
  project_name = "DGTX"
)

report_dir <- here::here(
  "results",
  "explore_data"
)
```

```{r Create the SplineOmics object, eval = TRUE}
# splineomics now contains the SplineOmics object.
splineomics <- SplineOmics::create_splineomics(
  data = data,
  meta = meta,
  report_info = report_info,
  condition = "Condition", # Column of meta that contains the levels.
  meta_batch_column = "Batch" # For batch effect removal in the plots
)

# Special print.SplineOmics function leads to selective printing
print(splineomics)
```

```{r Run EDA function, eval = FALSE}
plots <- SplineOmics::explore_data(
  splineomics = splineomics, # SplineOmics object
  report_dir = report_dir
)
```

[Here](https://drive.google.com/uc?id=1h4aOSwxUHAp5KAoAYVOlL_IRotV4EtY0)
you can see the HTML report of the explore_data() function with the NOT
batch-corrected data, and
[here](https://drive.google.com/uc?id=1op3oCt7eNhrUaofIkAyNJrSqpschU78b)
the report for the batch-corrected data.

# Run limma spline analysis

In this example, we are skipping finding the best hyperparameters with
the screen_limma_hyperparams() function, because we already have a clear
idea of what do use.

Lets define our parameters and put them into the `SplineOmics` object:

```{r Update the SplineOmics object, eval = TRUE}
splineomics <- SplineOmics::update_splineomics(
  splineomics = splineomics,
  # Reactor as mixed effect.
  design = "~ 1 + Condition*Time + Batch + (1|Reactor)", 
  mode = "integrated", # means limma uses the full data for each condition.
  spline_params = list(
    spline_type = c("n"), # natural cubic splines (take these if unsure)
    dof = c(2L) # If you are unsure about which dof, start with 2 and increase
  )
)
```

Run the `run_limma_splines()` function with the updated SplineOmics
object:

```{r limma spline analysis, eval = TRUE}
splineomics <- SplineOmics::run_limma_splines(
  splineomics = splineomics
)
```

# Build limma report

The topTables of all three limma result categories can be used to
generate p-value histograms an volcano plots.

```{r build limma report, eval = FALSE}
report_dir <- here::here(
  "results",
  "create_limma_reports"
)

plots <- SplineOmics::create_limma_report(
  splineomics = splineomics,
  report_dir = report_dir
)
```

You can view the generated analysis report of the create_limma_report
function
[here](https://drive.google.com/uc?id=1jgE8d5Gs12xnMATaitXxp28j2nAKCX1Y).

# Cluster the hits (significant features)

After we obtained the limma spline results, we can cluster the hits
based on their temporal pattern (their spline shape). We define what a
hit is by setting an adj. p-value threshold for every level. Hits are
features (e.g. proteins) that have an adj. p-value below the threshold.
Hierarchical clustering is used to place every hit in one of as many
clusters as we have specified for that specific level.

```{r cluster the hits, eval = FALSE}
adj_pthresholds <- c( # 0.05 for both levels
  0.05, # exponential
  0.05 # stationary
)

clusters <- c(
  2L, # 6 clusters for the exponential phase level
  2L # 3 clusters for the stationary phase level
)

report_dir <- here::here(
  "results",
  "clustering_reports"
)

# treatment_labels allows to place vertical dashed lines into the plots, that
# indicate a treatment, such as "temp shift" in this experiment. For each level,
# the treatment can be specified individually. For the "double spline plots", 
# where two levels are combined into one plot, treatment lines can also be 
# defined. The correct fieldname for those is {first_level}_{second_level}, 
# with first_level being the level of the two occuring first in the respective
# meta column, and second_level the one that occurs after. 
# Here, we don't want a treatment line for the constant level, since no 
# treatment was applied. To achieve that, we simply set it to NA or leave it
# out. 
treatment_labels = list(
  # constant = NA, 
  tshifted = "temp shift",  
  constant_tshifted = "temp shift" 
  )

# treatment_timepoints allows to specify the timepoint, at which the vertical
# dashed treatment line is placed. Again, for the constant level, we don't
# have a treatment, so we also do not specify a timepoint.
treatment_timepoints = list(
  # constant = NA,
  tshifted = 146,  
  constant_tshifted = 146
  )

plot_info <- list( # For the spline plots
  y_axis_label = "Glycan fractional abundance",
  time_unit = "hours", # our measurements were in minutes
  treatment_labels = treatment_labels,
  treatment_timepoints = treatment_timepoints
)

# Those are not genes for this glycan analysis here, but this argument expects
# the feature names (which usually are gene names, which is why it is called 
# like this.
genes <- rownames(data)

plot_options <- list(
  # When meta_replicate_column is not there, all datapoints are blue.
  meta_replicate_column = "Reactor", # Colors the data points based on Reactor
  cluster_heatmap_columns = FALSE # Per default FALSE, just for demonstration
)

clustering_results <- SplineOmics::cluster_hits(
  splineomics = splineomics,
  adj_pthresholds = adj_pthresholds,
  clusters = clusters,
  genes = genes,
  plot_info = plot_info,
  plot_options = plot_options,
  report_dir = report_dir,
  adj_pthresh_avrg_diff_conditions = 0.05,
  adj_pthresh_interaction_condition_time = 0.05
)
```

You can view the generated analysis report of the cluster_hits function
[here](https://drive.google.com/uc?id=1HXKoyOniO6-tDC-uCiGgpaQcjUQ9-zyl).

As discussed before, there are three limma result categories. The
cluster_hits() report shows the results of all three, if they are
present (category 2 and 3 can only be generated when the design formula
contains an interaction effect).

# Session Info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
