---
title: "limma_result_categories"
author: "Thomas Rauter"
date: "2025-01-21"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{limma_result_categories}
  %\VignetteEncoding{UTF-8}
---

# limma Result Categories

limma analysis results can be divided into three categories, which are defined in this document:

1.  **Time Effect**: This category focuses on the changes of the feature (e.g. protein) value that occur over time within a single condition.

2.  **Average Difference Between Conditions**: This category compares the average feature values for levels within a condition, regardless of time.

3.  **Interaction Between Condition and Time**: This category examines the interaction between time and condition. It identifies features whose value changes differently over time depending on the condition.

------------------------------------------------------------------------

### Legend:

-   **A hit** is a feature (e.g. a protein) that is significantly changed over time.
-   **Levels** are the different factors of a condition of the experiment. For example, bioreactor phase is a condition, and exponential and stationary are levels within that condition.

## Category 1 (time effect)

Temporal pattern within level for a given feature → Hit

#### Example of a Hit

A hit is a feature that shows a clear temporal pattern over time.

```{r echo=FALSE, fig.align='center', fig.height=4, fig.width=6, fig.cap="Clear temporal pattern over time"}
x <- seq(1, 10, length.out = 500) # More data points for smoother curve
y <- 10 + 5 * (sin(x) + 1) / 2 # Scale sin(x) to range [10, 15]

plot(x, y,
    type = "l", lwd = 2, col = "blue", ylim = c(10, 15),
    ylab = "Feature Value", xlab = "Time", main = "Category 1: Hit"
)
```

#### Example of No Hit

```{r echo=FALSE, fig.align='center', fig.height=4, fig.width=6, fig.cap="No clear temporal pattern"}
plot(1:10, rep(5, 10), type = "l", lwd = 2, col = "red", ylab = "Feature Value", xlab = "Time", main = "Category 1: No Hit")
```

## Category 2 (average difference conditions)

Overall mean difference between levels for a given feature → Hit

#### Example of a Hit

```{r echo=FALSE, fig.align='center', fig.height=4, fig.width=6, fig.cap="No clear temporal pattern for both levels but overall mean difference between them"}
x <- seq(1, 10, length.out = 500) # More data points for smoother curves
y1 <- 12 + 0.1 * sin(x / 2) # Slightly wavy, almost horizontal line for Level 1
y2 <- 13 + 0.1 * sin(x / 2) # Slightly wavy, almost horizontal line for Level 2

# Adjust margins to create space for the legend
par(mar = c(5, 4, 4, 8)) # Increase the right margin

plot(x, y1,
    type = "l", lwd = 2, col = "blue", ylim = c(10, 15),
    ylab = "Feature Value", xlab = "Time", main = "Category 2: Hit"
)
lines(x, y2, lwd = 2, col = "red")

# Add legend outside the plotting area
legend("topright",
    inset = c(-0.4, 0), legend = c("Level 1", "Level 2"),
    col = c("blue", "red"), lty = 1, lwd = 2, xpd = TRUE
)
```

#### Example of No Hit

```{r echo=FALSE, fig.align='center', fig.height=4, fig.width=6, fig.cap="Clear temporal pattern for both levels but no overall mean difference in feature value between them."}
x <- seq(1, 10, length.out = 100) # More data points for smooth lines
y1 <- 10 + 5 * (sin(x) + 1) / 2 # Scale sin(x) to range [10, 15]
y2 <- 10 + 5 * (cos(x) + 1) / 2 # Scale cos(x) to range [10, 15]

# Adjusting margins to create space for the legend
par(mar = c(5, 4, 4, 8)) # Increase the right margin

plot(x, y1,
    type = "l", lwd = 2, col = "blue", ylim = c(10, 15),
    ylab = "Feature Value", xlab = "Time", main = "Category 2: No Hit"
)
lines(x, y2, lwd = 2, col = "red")

# Add legend outside the plotting area
legend("topright",
    inset = c(-0.4, 0), legend = c("Level 1", "Level 2"),
    col = c("blue", "red"), lty = 1, lwd = 2, xpd = TRUE
)
```

## Category 3 (interaction condition & time)

Treatment interacting with time for a feature (time effect changing with treatment, the feature must have different temporal patterns in both conditions/levels) → Hit

#### Examples of Hits

Different temporal patterns are observed for each level --\> hit in category 3.

```{r echo=FALSE, fig.align='center', fig.height=4, fig.width=6, fig.cap="Different temporal patterns of the feature for both levels."}
x <- seq(1, 10, length.out = 100) # More data points for smooth lines
y1 <- 10 + 5 * (sin(x) + 1) / 2 # Scale sin(x) to range [10, 15]
y2 <- 10 + 5 * (cos(x) + 1) / 2 # Scale cos(x) to range [10, 15]

# Adjusting margins to create space for the legend
par(mar = c(5, 4, 4, 8)) # Increase the right margin

plot(x, y1,
    type = "l", lwd = 2, col = "blue", ylim = c(10, 15),
    ylab = "Feature Value", xlab = "Time", main = "Category 3: Hit"
)
lines(x, y2, lwd = 2, col = "red")

# Add legend outside the plotting area
legend("topright",
    inset = c(-0.4, 0), legend = c("Level 1", "Level 2"),
    col = c("blue", "red"), lty = 1, lwd = 2, xpd = TRUE
)
```

```{r echo=FALSE, fig.align='center', fig.height=4, fig.width=6, fig.cap="Different temporal patterns of the feature for both levels."}
x <- seq(1, 10, length.out = 100) # More data points for smooth lines
y1 <- 0.25 + 0.75 * (x / max(x)) # Straight upward trend
y2 <- 0.75 - 0.75 * (x / max(x)) # Straight downward trend

# Adjust margins to create space for the legend
par(mar = c(5, 4, 4, 8)) # Increase the right margin

plot(x, y1,
    type = "l", lwd = 2, col = "blue", ylim = c(0, 1),
    ylab = "Feature Value", xlab = "Time", main = "Category 3: Hit"
)
lines(x, y2, lwd = 2, col = "red")

# Add legend outside the plotting area
legend("topright",
    inset = c(-0.4, 0), legend = c("Level 1", "Level 2"),
    col = c("blue", "red"), lty = 1, lwd = 2, xpd = TRUE
)
```

#### Example of No Hit

```{r echo=FALSE, fig.align='center', fig.height=4, fig.width=6, fig.cap="Overall the same temporal pattern of a feature for both levels."}
x <- seq(1, 10, length.out = 100) # More data points for smooth lines
y1 <- 0.75 - 0.5 * (x / max(x)) # First downward trend
y2 <- 0.65 - 0.4 * (x / max(x)) # Second downward trend with a slightly different slope

# Adjust margins to create space for the legend
par(mar = c(5, 4, 4, 8)) # Increase the right margin

plot(x, y1,
    type = "l", lwd = 2, col = "red", ylim = c(0, 1),
    ylab = "Feature Value", xlab = "Time", main = "Category 3: No Hit"
)
lines(x, y2, lwd = 2, col = "blue")

# Add legend outside the plotting area
legend("topright",
    inset = c(-0.4, 0), legend = c("Line 1", "Line 2"),
    col = c("red", "blue"), lty = 1, lwd = 2, xpd = TRUE
)
```

# Session Info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
