---
title: "PVC_analysis"
author: "Thomas Rauter"
date: "23 April, 2025"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PVC_analysis}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# About this tutorial

This tutorial demonstrates the **excursion-analysis** capabilities of the `SplineOmics` package by walking through a complete, real-world phosphoproteomics example from
start to finish.

In the context of time series data, an **excursion** refers to a **transient deviation from a baseline** — a concept borrowed from signal processing. These short-term, local shifts can reveal biologically meaningful events that are not captured by global trends.

Details of how excursions are detected are described in the section on the **PVC test**, a method specifically designed to identify such local patterns.

While the core strength of `SplineOmics` lies in modeling **global temporal patterns** using smoothing splines, many biologically important phenomena present as **localized events**. These may not be well captured by standard spline fits due to their smooth and global nature.

For example, in phosphoproteomics data, you might observe a **sharp, temporary phosphorylation spike** at a specific timepoint following a stimulus — a pattern that returns to baseline shortly afterward. Such a transient peak can indicate a key signaling activation event that is **functionally critical**, but would be smoothed out or diluted in traditional spline-based analyses.

To handle these scenarios, we developed the **PVC test** (described below), which is specifically designed to detect such **sharp, local excursions** in time-resolved omics data.

## Note 1

The documentation of all the **SplineOmics** package functions can be
viewed [here](https://csbg.github.io/SplineOmics/reference)


## Note 2 

This vignette focuses on the **excursion-analysis**. The general functionalities
of `SplineOmics` are explained in the **get-started** vignette.



# The PVC Test

**PVC** stands for **P**eak, **V**alley, and **C**liff. The PVC test is a method to identify these distinct local patterns (called excursions) in time series data using a compound contrast approach within the **limma** framework.

In this context:

- A **peak** occurs when a timepoint is significantly **higher** than both its immediate neighbors.
- A **valley** occurs when a timepoint is significantly **lower** than both its neighbors.
- A **cliff** describes a sharp directional shift: when one neighbor is similar in value, but the other is **significantly different** (higher or lower).


## Method

The PVC analysis is implemented as a **rolling, compound contrast** applied across the time series. Specifically, a window of three consecutive timepoints is evaluated at a time:  
`(Tᵢ₋₁, Tᵢ, Tᵢ₊₁)`  
where `Tᵢ` is the center of the window.

The first and last timepoints are excluded from this analysis, since they lack both left and right neighbors.

For each triplet, a contrast is defined as:

2*Tᵢ - Tᵢ₋₁ - Tᵢ₊₁ 

This contrast measures whether the center point `Tᵢ` deviates from both its neighbors in the same direction:

- If `Tᵢ` is **greater than both neighbors** → large positive contrast → indicates a **peak**
- If `Tᵢ` is **lower than both neighbors** → large negative contrast → indicates a **valley**
- If `Tᵢ` lies **between** its neighbors → effects cancel → contrast is near zero → **no signal**

This contrast is tested using `limma`, and the result is a p-value that indicates whether the observed pattern is statistically significant for each feature (e.g., gene, phosphosite).


## Multiple Testing Correction

Because the test is performed independently at each center timepoint across all features, **multiple testing correction** is applied at the **timepoint level**. This ensures that significance is not inflated due to repeated testing across features.

Importantly, PVC does **not** assess whether an entire feature contains a full peak/valley/cliff pattern over time — it evaluates **each timepoint independently** for such local patterns.



# Details about the dataset

The example dataset used here involves a **time-series phosphoproteomics experiment**,
where CHO (chinese hamster ovary) cells were cultivated in three bioreactors
(three biological replicates). The experiment includes the following
setup:

-   Samples were taken both during the **exponential** and **stationary
    growth phases**.
-   Samples were collected in triplicates from each reactor at defined
    timepoints relative to cell feeding:
    -   60 minutes before feeding
    -   15, 60, 90, 120, and 240 minutes after feeding

Note that the dataset was truncated to 1000 rows for file size reasons and the
annotation info, such as gene name, was randomly shuffled.


# Load the packages

```{r setup, eval = TRUE}
library(SplineOmics)
library(here) # For managing filepaths
library(dplyr) # For data manipulation
library(knitr) # For Showing the head of the data and the meta tables.
```

# Load the files

In this example, the proteomics_data.rds file contains the numeric
values (the intensities) and also the feature descriptions, such as gene
and protein name (= annotation part). Usually, you would load the data
from for example an Excel file, but the .rds file is more compressed,
which is the reason this format was chosen here to limit the size of the
SplineOmics package.

The file meta.xlsx contains the meta information, which are the
descriptions of the columns of the numeric values of data.

(These example files are part of the package and don't have to be
present on your system).

Please note that this dataset is an actual experimental dataset, but the
annotation information, such as gene names, has been removed since it
was not yet published at the time of making the SplineOmics package
public. Instead, the dataset includes randomly generated gene symbols
and gene names corresponding to Cricetulus griseus (Chinese Hamster) for
each row. This is intended to demonstrate the functionality of the
package.

The left part of data contains the numeric values, and the right part
the annotation info, which can be copied in a separate dataframe, as
shown below.

```{r load the files}
data <- readRDS(xzfile(system.file(
  "extdata",
  "phosphoproteomics_data.rds.xz",
  package = "SplineOmics"
)))

meta <- read.csv(
  system.file(
    "extdata",
    "phosphoproteomics_meta.csv",
    package = "SplineOmics"
  ),
  stringsAsFactors = FALSE
)

# Extract the annotation part from the dataframe.
first_na_col <- which(is.na(data[1, ]))[1]
annotation <- data |> dplyr::select((first_na_col + 1):ncol(data)) 
```

## Show top rows of data

```{r Show data head, eval = TRUE}
kable(
  head(data),
  format = "markdown"
  )
```

## Show top rows of meta

```{r Show meta head, eval = TRUE}
kable(
  head(meta),
  format = "markdown"
  )
```

## Show top rows of annotation

```{r Show annotation head, eval = TRUE}
kable(
  head(annotation),
  format = "markdown"
  )
```


## Bring the Inputs into the Standardized Format

```{r process inputs, eval = TRUE}
data <- SplineOmics::extract_data(
  # The dataframe with the numbers on the left and info on the right.
  data = data,
  # Use this annotation column for the feature names.
  feature_name_columns = c("T..Gene"),
  use_row_index = TRUE,   # makes the feature names unique with row index
  top_row = 1,
  bottom_row = 1000,
  right_col = 36,
  left_col = 1
)
```

## SplineOmics Object

```{r Define report info, eval = TRUE}
# Those fields are mandatory, because we believe that when such a report is
# opened after half a year, those infos can be very helpful.
report_info <- list(
  omics_data_type = "PPTX",
  data_description = "Phosphoproteomics data of CHO cells",
  data_collection_date = "February 2024",
  analyst_name = "Thomas Rauter",
  contact_info = "thomas.rauter@plus.ac.at",
  project_name = "DGTX"
)
```

```{r Create the SplineOmics object, eval = TRUE}
# splineomics now contains the SplineOmics object.
splineomics <- SplineOmics::create_splineomics(
  data = data,
  meta = meta,
  annotation = annotation,
  report_info = report_info,
  condition = "Phase", # Column of meta that contains the levels.
  meta_batch_column = "Reactor", # For batch effect removal
)
```


# Run the PVC-analysis

```{r Define plot_info, eval = TRUE}
plot_info <- list(
  y_axis_label = "log2 intensity",
  time_unit = "min",
  treatment_labels = list(
    Exponential = "temp shift",
    Stationary = "temp shift"
    ),
  treatment_timepoints = list(
    Exponential = 146,
    Stationary = 146
    )
)
```

```{r run find_pvc function, eval = TRUE}
# Check out the documentation of the function under the Reference tab.
excursion_plots <- SplineOmics::find_pvc(
  splineomics = splineomics,
  alphas = 0.025,
  plot_info = plot_info,
  report_dir = withr::local_tempdir()
)
```

The terminal output of the function gives information about the total amount
of excursions found in each level, and also a breakdown by timepoint and type
of excursion (p = peak, v = valley, b = bottom of cliff, t = top of cliff,
always for the respective timepoint that is mentioned in that row).
[Here](https://csbg.github.io/SplineOmics_html_reports/pvc_report_PPTX.html)
you can see the resulting HTML report. 

# Session Info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```
