---
title: "SplineOmics-Tutorial"
author: "Thomas Rauter"
date: "10 June, 2024"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SplineOmics-Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# About this tutorial

This tutorial intends to showcase and explain the capabilities of the SplineOmics
package by walking through a real and complete example, from start to the end.
The example is a time-series proteomics experiment, in which CHO cells were cultivated in three bioreactors (three biological replicates). Both in the exponential and stationary growth phase, samples were taken from each reactor in triplicates at defined timepoints relative to a feeding of the cells (60 min before feeding, and 15, 60, 90, 120, and 240 min after). The goal of this analysis is to find out which of the 7162 cellular proteins show a significant change over time after the CHO cells were fed. Further, the hits (proteins with a significant change over time) are clustered based on their temporal pattern, and a gene set enrichment analysis is performed with each cluster to see if there are processes which are up- or downregulated over time after the feeding. 

# Load the packages

```{r setup}
library(SplineOmics)
library(readxl)
```

# Load the files

```{r load the files}
data <- read_excel(system.file("extdata", "data.xlsx", package = "SplineOmics"))
annotation <- read_excel(system.file("extdata", "annotation.xlsx", package = "SplineOmics"))
meta <- read_excel(system.file("extdata", "meta.xlsx", package = "SplineOmics"))

print(data)
print(annotation)
print(meta)
```

# Perform EDA (exploratory data analysis)

The first step in analysing data is usually EDA. EDA involves summarizing the 
main characteristics of the data, often using plots. This can involve densitiy 
distributions, boxplots, PCA, correlation heatmaps, and more. This process can
be carried out by using the package function explore_data(). The neccessary arguments are the data matrix (data), the meta table (meta), the name of the column that contains the levels of the experiment (condition) (the levels here are Exponential and Stationary), and the report_info list, that contains general info about the analysis. Optional arguments are 

```{r Perform EDA}
report_info <- list(
  omics_data_type = "PTX",
  data_description = "Proteomics data of CHO cells",
  data_collection_date = "February 2024",
  analyst_name = "Thomas Rauter",
  contact_info = "thomas.rauter@plus.ac.at",
  project_name = "DGTX")

condition <- "Phase"
meta_batch_column <- "Reactor"
report_dir <- here::here("results", "explore_data")

plots <- explore_data(data = data,
                      meta = meta,
                      condition = condition,
                      report_info = report_info,
                      meta_batch_column = meta_batch_column,
                      report_dir = report_dir)

```

You can view the detailed analysis report [here](../inst/reports/report.html).